---
title: "R Programming Course Pack"
author: " "
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

# Course Overview

This course introduces students to **R programming** from first principles and builds toward **intermediate data analysis and visualization skills**.

**Audience:** First-year university students with mixed prior experience  
**Duration:** 10--12 weeks  
**Format:** Weekly lecture + lab + assignment

# Learning Outcomes

By the end of this course, students should be able to:

- Use R and RStudio confidently
- Create and manipulate variables, vectors, and data frames
- Import, explore, and clean datasets
- Write simple programs using conditionals, loops, and functions
- Use `dplyr` for data wrangling
- Create plots using `ggplot2`
- Carry out basic statistical modelling in R
- Prepare a simple reproducible report in R Markdown

# How to Use This Course Pack

Each week includes:

- Learning objectives
- Lecture notes
- Live coding examples
- Practice exercises
- Assignment ideas

# Week 1: Introduction to R and RStudio

## Objectives

- Understand what R is used for
- Become familiar with the RStudio interface
- Run basic commands
- Create and use variables

## Lecture Notes

R is a programming language widely used for statistics, data analysis, and visualization.  
RStudio is an integrated development environment that makes it easier to write and run R code.

Students should learn:
- Console vs script
- Environment
- Plots pane
- Files pane

## Live Coding Examples

```{r}
# Basic arithmetic
2 + 3
10 - 4
5 * 6
12 / 3

# Assigning variables
x <- 10
y <- 5

x + y
x * y
```

## Practice Exercises

1. Create a variable `a` equal to 7.
2. Create a variable `b` equal to 3.
3. Compute:
   - `a + b`
   - `a - b`
   - `a * b`
   - `a / b`

## Assignment

Write a short R script that calculates the total cost of 3 items and then adds tax.

# Week 2: Data Types and Vectors

## Objectives

- Understand common R data types
- Create vectors
- Use basic summary functions

## Lecture Notes

Common data types in R include:
- numeric
- character
- logical

A vector is a collection of values of the same type.

## Live Coding Examples

```{r}
# Data types
num <- 10
word <- "Hello"
flag <- TRUE

class(num)
class(word)
class(flag)

# Vectors
scores <- c(80, 90, 75, 88)

mean(scores)
sum(scores)
length(scores)
max(scores)
min(scores)
```

## Practice Exercises

1. Create a numeric vector with 5 values.
2. Find its mean.
3. Find its maximum and minimum.
4. Create a character vector with the names of 4 students.

## Assignment

Store 10 student scores in a vector and compute the mean, median, minimum, and maximum.

# Week 3: Data Frames and Importing Data

## Objectives

- Understand tabular data in R
- Create a data frame
- Import a CSV file
- Explore a dataset

## Lecture Notes

A data frame stores data in rows and columns.  
Each column can represent a variable, and each row can represent an observation.

Useful exploration functions:
- `head()`
- `str()`
- `summary()`
- `names()`

## Live Coding Examples

```{r}
students <- data.frame(
  name = c("Alice", "Bob", "Chidi", "Dana"),
  age = c(18, 19, 18, 20),
  score = c(85, 90, 78, 92)
)

head(students)
str(students)
summary(students)
names(students)
```

```{r eval=FALSE}
# Importing a CSV file
df <- read.csv("data.csv")
head(df)
```

## Practice Exercises

1. Create a data frame with 3 columns and 5 rows.
2. Display its structure.
3. Show the first few rows.
4. Find the names of the columns.

## Assignment

Import a CSV file into R and write 5 observations about the dataset.

# Week 4: Subsetting and Filtering Data

## Objectives

- Extract rows and columns
- Filter data using logical conditions
- Access variables in a data frame

## Lecture Notes

Subsetting allows students to isolate the data they need.

Examples:
- Select rows
- Select columns
- Filter based on a condition

## Live Coding Examples

```{r}
students
students$score
students[, c("name", "score")]
students[students$score > 80, ]
students[students$age == 18, ]
```

## Practice Exercises

1. Select only the `name` column.
2. Select rows where score is above 85.
3. Select rows where age is 18.
4. Select only the `name` and `age` columns.

## Assignment

Using a small student dataset, identify all students who passed and all students who scored above 90.

# Week 5: Control Flow

## Objectives

- Use `if`, `else if`, and `else`
- Understand the role of logical conditions
- Introduce loops

## Lecture Notes

Control flow allows code to make decisions.

## Live Coding Examples

```{r}
score <- 75

if (score >= 50) {
  print("Pass")
} else {
  print("Fail")
}
```

```{r}
score <- 85

if (score >= 80) {
  print("A")
} else if (score >= 70) {
  print("B")
} else if (score >= 60) {
  print("C")
} else {
  print("Needs Improvement")
}
```

```{r}
for (i in 1:5) {
  print(i)
}
```

## Practice Exercises

1. Write code that prints `"Even"` if a number is even and `"Odd"` otherwise.
2. Write a loop that prints the numbers from 1 to 10.
3. Write code that assigns a letter grade based on a score.

## Assignment

Write a grading program that takes a score and returns a category such as A, B, C, D, or F.

# Week 6: Writing Functions

## Objectives

- Understand why functions are useful
- Write simple custom functions
- Use inputs and outputs

## Lecture Notes

Functions help avoid repetition and make code more organized.

## Live Coding Examples

```{r}
grade_fun <- function(score) {
  if (score >= 50) {
    return("Pass")
  } else {
    return("Fail")
  }
}

grade_fun(70)
grade_fun(42)
```

```{r}
add_two_numbers <- function(x, y) {
  x + y
}

add_two_numbers(3, 5)
```

## Practice Exercises

1. Write a function that doubles a number.
2. Write a function that calculates the area of a rectangle.
3. Write a function that classifies a score as pass or fail.

## Assignment

Create at least 3 useful functions and test them with multiple inputs.

# Week 7: Data Wrangling with dplyr

## Objectives

- Use `filter()`, `select()`, `mutate()`, and `arrange()`
- Understand the pipe operator `%>%`
- Begin modern data wrangling workflows

## Lecture Notes

The `dplyr` package makes data manipulation easier and more readable.

## Live Coding Examples

```{r}
library(dplyr)

students %>%
  filter(score > 80)

students %>%
  select(name, score)

students %>%
  mutate(pass = score >= 50)

students %>%
  arrange(score)
```

## Practice Exercises

1. Filter students with score greater than 85.
2. Select only `name` and `score`.
3. Create a new variable called `pass`.
4. Arrange the data by score from lowest to highest.

## Assignment

Use `dplyr` to clean and summarize a small dataset.

# Week 8: Grouping and Summarising Data

## Objectives

- Use `group_by()` and `summarise()`
- Compute grouped statistics
- Compare categories

## Lecture Notes

Grouping is useful when we want summaries by category, such as average scores by group.

## Live Coding Examples

```{r}
students2 <- data.frame(
  name = c("Alice", "Bob", "Chidi", "Dana", "Eva", "Farah"),
  group = c("A", "A", "B", "B", "A", "B"),
  score = c(85, 90, 78, 92, 88, 81)
)

students2 %>%
  group_by(group) %>%
  summarise(
    avg_score = mean(score),
    max_score = max(score),
    n = n()
  )
```

## Practice Exercises

1. Group data by one variable.
2. Compute average score by group.
3. Count how many students are in each group.

## Assignment

Using a grouped dataset, calculate summary statistics and interpret the results in words.

# Week 9: Data Visualization with ggplot2

## Objectives

- Understand the grammar of graphics
- Create histograms, scatterplots, and bar charts
- Label plots clearly

## Lecture Notes

Good visualizations help communicate results clearly.

## Live Coding Examples

```{r}
library(ggplot2)

ggplot(students, aes(x = score)) +
  geom_histogram(binwidth = 5)
```

```{r}
ggplot(students, aes(x = age, y = score)) +
  geom_point()
```

```{r}
ggplot(students, aes(x = name, y = score)) +
  geom_col()
```

## Practice Exercises

1. Create a histogram of scores.
2. Create a scatterplot of age vs score.
3. Create a bar chart of names and scores.

## Assignment

Create 3 plots from a dataset and write 2--3 sentences interpreting each one.

# Week 10: Improving Plots

## Objectives

- Customize titles and labels
- Use themes
- Create faceted plots
- Improve readability

## Lecture Notes

Students should learn that a good plot is not only correct but also clear and visually effective.

## Live Coding Examples

```{r}
ggplot(students2, aes(x = group, y = score)) +
  geom_boxplot() +
  labs(
    title = "Scores by Group",
    x = "Group",
    y = "Score"
  )
```

```{r}
ggplot(students2, aes(x = score)) +
  geom_histogram(binwidth = 5) +
  facet_wrap(~ group)
```

## Practice Exercises

1. Add a title and axis labels to a plot.
2. Create a boxplot by group.
3. Create a faceted histogram.

## Assignment

Redesign a basic plot to make it publication-ready for a class audience.

# Week 11: Iteration and Repetition

## Objectives

- Review loops
- Introduce the `apply` family
- Practice repeated operations

## Lecture Notes

Iteration is useful when we need to repeat the same process many times.

## Live Coding Examples

```{r}
for (i in 1:5) {
  print(i^2)
}
```

```{r}
mat <- matrix(1:9, nrow = 3)
mat

apply(mat, 2, sum)
apply(mat, 1, mean)
```

## Practice Exercises

1. Use a loop to print the squares of numbers 1 to 10.
2. Create a matrix and compute column sums.
3. Use `apply()` to compute row means.

## Assignment

Compare solving a repeated task using a loop and using `apply()`.

# Week 12: Introduction to Statistical Modelling

## Objectives

- Fit a simple linear regression model
- Interpret model output
- Connect data analysis and modelling

## Lecture Notes

A linear model studies the relationship between a response variable and one or more predictors.

## Live Coding Examples

```{r}
study_data <- data.frame(
  study_hours = c(2, 3, 4, 5, 6, 7, 8),
  score = c(55, 60, 65, 70, 74, 80, 85)
)

model <- lm(score ~ study_hours, data = study_data)
summary(model)
```

```{r}
plot(study_data$study_hours, study_data$score)
abline(model)
```

## Practice Exercises

1. Fit a simple linear model.
2. Identify the slope and intercept.
3. Explain what the slope means in context.

## Assignment

Fit a linear model to a small dataset and explain the meaning of the results in plain language.

# Assessment Structure

## Suggested Breakdown

- Weekly exercises: 20%
- Lab participation: 10%
- Midterm coding task: 20%
- Final project: 30%
- Final quiz or test: 20%

# Final Project

Students should complete a small end-of-course project where they:

1. Import a dataset
2. Explore and clean the data
3. Compute summary statistics
4. Create at least 2 useful plots
5. Fit one simple model or perform one structured analysis
6. Write a short interpretation of their findings

# Differentiation for Mixed-Ability Students

## Support for Beginners

- Step-by-step worksheets
- Guided examples
- Small group coding practice
- Glossary of basic terms

## Extension for More Experienced Students

- Challenge exercises each week
- Open-ended mini projects
- Optional use of additional packages
- Function-writing and data-cleaning extensions

# Bonus Unit: Introduction to R Markdown

## Objectives

- Understand reproducible reports
- Combine text, code, and output in one document
- Knit a simple report

## Example

```{r eval=FALSE}
---
title: "My First Report"
output: html_document
---

# Introduction

This is my first R Markdown document.

```{r}
x <- 1:10
mean(x)
```
```

# Suggested Datasets

You may use:
- Student performance data
- Sales data
- Weather data
- Sports statistics
- Health or survey data

# Closing Note

This course pack is designed to move students from beginner to intermediate level in a structured and supportive way. It can be adapted for a semester course, a bootcamp, or a workshop series.
